Simulating High-Dimensional Multivariate Data using the \pkg{bigsimr} Package

### 2021-06-22 000-introduction.Rmd
removed duplicated ref
Revise for JOSS submission. Address Tom's comments

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

>>>>

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependence perfectly. However, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

### 2021-06-22 010-background.Rmd
Revise for JOSS submission. Address Tom's comments

\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined by \@ref(eq:inverseCDF) when the variables are discrete.

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
\label{eq:inverseCDF}
\end{equation}

@@

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix $R_P$. Since the marginals are standardized to have unit variance, however, $R_P$ is a Pearson correlation matrix. If $F_R$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_R$ is defined through

@@

Here the copula function $C_R$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz. $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

@@

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.  In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

>>>>

\noindent where $U$ is a uniform random variable on $(0,1)$ and $F^{-1}_1, F^{-1}_2$ are the inverse CDFs of $X_1$ and $X_2$, respectively, defined as

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}
\label{eq:inverseCDF}
\end{equation}

when the variables are discrete.

@@

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, . This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix $R_P$. Since the marginals are standardized to have unit variance, this $R_P$ is also Pearson correlation matrix. If $F_R$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_R$ is defined through

@@

where $\Phi(\cdot)$ is the standard normal CDF.

Here, the copula function $C_R$ is the CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

Sklar's Theorem [@Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz. $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

@@

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.  In `R`, the `MASS` package's `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures and it fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released, which provides HD multivariate normal generation. However, lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vector generation via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

### 2021-06-22 020-simulation-algorithm.Rmd
Revise for JOSS submission. Address Tom's comments

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as:

@@

1. Simulate a random vector $\bf Z$ with $d$ independent and identically distributed (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.

@@

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\binom{d}{2}$ double integrals, as suggested by @XZ19.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@@

There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

@@

For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins. To this end, we implement the approximation scheme introduced by @XZ19. Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive. To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions [@muino2006gs].

*Step 2: Admissibility check and nearest correlation matrix computation.* Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above. If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing. As implemented the function fails to scale to HD. Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions (@qi2006quadratically). The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.

>>>>

The well-known NORTA algorithm [@Cario1997] simulates a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$. Specifically, NORTA algorithm proceeds as follows:

@@

1. Simulate a random vector $\bf Z$ with $d$ independent and identically distributed (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; {\bf Y} \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.

@@

where $\phi(\cdot)$ is the joint probability density function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\binom{d}{2}$ double integrals, as suggested by @XZ19.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix becomes non-positive definiteness. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@@

There is hope, however, as @GH02 also showed that replacing an non-positive definite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

@@

For the computationally expensive process to match Pearson correlations, we approximate Equation \@ref(eq:pearsonIntegralRelation) for all pairs of margins. To this end, we implement the approximation scheme introduced by [@XZ19]. Briefly, the many double integrals of the form in \@ref(eq:pearsonIntegralRelation) are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the approximation is computationally expensive. To remedy this, we further approximate discrete distributions by a continuous distribution using Generalized S-Distributions [@muino2006gs].

*Step 2: Admissibility check and nearest correlation matrix computation.* Once $R_X$ has been determined, we check admissibility of the adjusted correlation via the steps described above. If $R_X$ is not a valid correlation matrix, then we compute the nearest correlation matrix. Finding the nearest correlation matrix is a common statistical computing problem. The defacto function in `R` is `Matrix::nearPD`, an alternating projection algorithm due to @higham2002computing. As implemented, the function fails to scale to HD. Instead, we provide the quadratically-convergent algorithm based on the theory of strongly semi-smooth matrix functions [@qi2006quadratically]. The nearest correlation matrix problem can be written down as the following convex optimization problem:  $\mathrm{min} \quad \frac{1}{2} \Vert R_X - X \Vert^2, \quad \mathrm{s.t.} \quad X_{ii} = 1, \quad i = 1, \ldots , n, \quad X \in S_{+}^{n}$.

### 2021-06-22 040-monte-carlo-evaluation.Rmd
Revise for JOSS submission. Address Walt's comments.
Revise for JOSS submission. Address Tom's comments

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results. Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios. Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.

@@

Somewhat surprising, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching. In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization. This could mitigate some of the current algorithmic/implementation shortcomings. Despite this, Kendall matching is still feasible for most HD data sets. Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a 'close' (not nearest PD) to scale to even higher dimensions with some loss of accuracy.

>>>>

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results. Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios. Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for the negative binomial case.

@@

Somewhat surprisingly, Kendall estimation and nearest PD computation scale poorly compared to Spearman and, even, approximate Pearson matching. In our experience, Kendall computation times are sensitive to the number of cores, benefiting from multi-core parallelization. This could mitigate some of the current algorithmic/implementation shortcomings. Despite this, Kendall matching is still feasible for most HD data sets. Finally, we note that one could use our single-pass algorithm `cor_fastPD` to produce a 'close' (not nearest PD) to scale to even higher dimensions with some loss of accuracy.

### 2021-06-22 050-examples.Rmd
Revise for JOSS submission. Address Tom's comments

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of HD dependent count data. In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process. Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation occurs during biological processes [@Wang2009b]. And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction. See @Efron2012 for a detailed discussion with related methods and @Wu2012b, @Schissler2018; @Schissler2019 for applied examples. The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.

@@

As seen above $R_X$ is not strictly admissible in our scheme. This can be seen from the negative result from `bs$iscorrelation(R_X)`, which checks 1) positive-definiteness, 2) symmetry, and 3) 1s along the diagonal. However, the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations. With the inputs configured, `rvec` is executed to produce a synthetic RNA-seq data set:

>>>>

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of HD dependent count data. In an illustration of our proposed methodology, we seek to simulate RNA-sequencing data by producing simulated random vectors mimicking the observed data and its generating process. Modeling RNA-seq using multivariate probability distributions is natural as inter-gene correlation occurs during biological processes [@Wang2009b]. And yet, many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction. The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicating an estimated parametric structure, probability estimation, and evaluation of correlation estimation efficiency.

@@

As seen above, $R_X$ is not strictly admissible in our scheme. This can be seen from the negative result from `bs$iscorrelation(R_X)`, which checks if $R_X$ is positive-definite with 1's along the diagonal. However, the approximation is close with a maximum absolute error of `r max( abs(targets - approximates))` and average absolute error of `r mean(abs(targets - approximates))` across the `r choose(d,2)` correlations. With the inputs configured, `rvec` is executed to produce a synthetic RNA-seq data set:

### 2021-06-22 060-conclusion-discussion.Rmd
Revise for JOSS submission. Address Tom's comments

We developed a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package `bigsimr`, Julia package `Bigsimr` and Python package `bigsimr`. The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17]. The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad data analytic applications for modern, big-data statistical computing. For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure. Or our methods could be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.

@@

There are limitations to the methodology and implementation. We could only investigate selected multivariate distributions in our Monte Carlo studies. There may be instances that the methods do not perform well. Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure. Another issue is that for discrete distributions, we use continuous approximations when the support set is large. This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions. Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.

>>>>

We developed a general-purpose high-dimensional multivariate simulation algorithm and provide a user-friendly, high-performance `R` package `bigsimr`, Julia package `Bigsimr` and Python package `bigsimr`. The random vector generation method is inspired by NORTA [@Cario1997] and Gaussian copula-based approaches [@MB13, @BF17, @Xia17]. The major contributions of this work are methods and software for flexible, scalable simulation of HD multivariate probability distributions with broad data analytic applications for modern, big-data statistical computing. For example, one could simulate high-resolution time series data, such as those consistent with an auto-regressive moving average model exhibiting a specified Spearman structure. Our methods could also be used to simulate sparsely correlated data, as many HD methods assume, via specifying a *spiked correlation matrix*.

@@

There are limitations to the methodology and implementation. We could only investigate selected multivariate distributions in our Monte Carlo studies. There may be instances where the methods do not perform well. Along those lines, we expect that correlation values close to the boundary of the feasible region could result in algorithm failure. Another issue is that for discrete distributions, we use continuous approximations when the support set is large. This could limit the scalability/accuracy for particular discrete/mixed multivariate distributions. Our method would also benefit computationally from faster Kendall estimation and more finely tuned nearest PD calculation for non-admissible Kendall matrices.
